---
title: "HW2"
author: "Adrián Arribas and Rodrigo Milton Campos"
output: html_document
date: "2025-12-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
rm(list = ls())
```
# Homework 2

We choose "Exited" as the target variable for our study to determine whether a customer leaves the bank or not.

Here are the rest of the variables:

CustomerId: A unique numerical identification number for each customer.

CreditScore: A numerical value used to measure the customer's creditworthiness.

Gender: The gender of the customer (Male or Female).

Balance: The amount of money currently held in the customer's bank account.

NumOfProducts: The total number of different bank products or services used by the customer.

HasCrCard: A binary indicator showing whether the customer possesses a credit card.

IsActiveMember: A binary indicator representing whether the customer is considered an active member of the bank.

Age: The age of the customer.

Geography: The region where the customer is located (Texas, California, or Alabama).

EstimatedSalary: The projected annual salary of the customer.

Surname: The customer's last name.

Tenure: The number of years the customer has been a client of the bank.


We load all the necessary libraries
```{r, include = FALSE}
library(mice)
library(GGally)
library(caret)
library(randomForest)
library(dplyr)
```

Loading the dataset:
```{r, include=FALSE}
set.seed(123)
pre_data = read.csv("df_estados_bank.csv")
```

##Preprocessing
```{r}
summary(pre_data)
```

Remove ID's
```{r}
pre_data = pre_data[,4:15]
```

We also remove the surnames, as we will not be using them
```{r}
pre_data = pre_data[,-10]
```

Creating dummy variables for each state
```{r}
pre_data$Geography = as.factor(pre_data$Geography)
str(pre_data$Geography)
levels(pre_data$Geography)
```

```{r}
pre_data$Texas = ifelse(pre_data$Geography == "Texas", 1, 0)
pre_data$California = ifelse(pre_data$Geography == "California", 1, 0)
pre_data$Alabama = ifelse(pre_data$Geography == "Alabama", 1, 0)

head(pre_data[c("Geography", "Alabama", "California", "Texas")])
# Removing geography
pre_data = subset(pre_data, select = -c(Geography))
```

Checking NA's
```{r}
barplot(colMeans(is.na(pre_data)), las=2)
```
Cleaning NA's
```{r}
clean = mice(pre_data, method = "rf", m = 5)
data = complete(clean)
```

It is safe to assume a tenure of 100 years is not the intended value and probably a typo or a mistake
```{r}
data = data[data$Tenure < 100,]
```

Checking for correlation between variables
```{r}
data$Gender = as.factor(data$Gender)
levels(data$Gender)
data_num = data[,-2]
data_num$Gender = as.numeric(data$Gender) - 1 # 0 for female and 1 for male
```


```{r, echo=FALSE}
R = cor(data_num)
ggcorr(data_num, label = T)
```

Factor Conversion
```{r}
#Convert Gender to Factor
data$Gender = as.factor(data$Gender)

#Convert Exited to Factor
data$Exited = as.factor(data$Exited)

#Convert Numerical Categorical Variables to Factors
data$NumOfProducts = as.factor(data$NumOfProducts)
data$HasCrCard = as.factor(data$HasCrCard)
data$IsActiveMember = as.factor(data$IsActiveMember)
data$Tenure = as.factor(data$Tenure)

#Check the final structure
str(data)
```

We visualize the target variable
```{r}
ggplot(data, aes(x = Age, fill = Exited)) +
  geom_density() +
  labs(title = "Exited Distribution by Age",
       x = "Age",
       fill = "Exited") +
  theme_minimal()
```

Although the plot of those that exited seems to be shifted towards a higher age, there is not much of a difference.

```{r}
ggplot(data, aes(x = NumOfProducts, fill = Exited)) +
  geom_bar() +
  labs(title = "Exited Distribution by Products",
       x = "Number of Products",
       fill = "Exited") +
  theme_minimal()
```

Even though the count is rather small, there is a much bigger proportion of customers that exited when looking at those with a higher number of products used.

```{r}
ggplot(data, aes(x = Balance, fill = Exited)) +
  geom_histogram() +
  labs(title = "Exited Distribution by Balance",
       x = "Balance",
       fill = "Exited") +
  theme_minimal()
```

Yields no clear conclusion as to how the balance affects the target variable.

##Classification

Data partition and display check
```{r}
train_index = createDataPartition(data$Exited, p = 0.7, list = FALSE)

# Split the data
train_data = data[train_index, ]
test_data  = data[-train_index, ]

# Display
prop.table(table(train_data$Exited))
prop.table(table(test_data$Exited))
```

###Random Forest
Training the random forest
```{r}
rf_model = randomForest(
  Exited ~ ., 
  data = train_data, 
  ntree = 500,
  importance = TRUE
)

# Print the model summary
print(rf_model)
```
Evaluating the random forest
```{r}
rf_predictions <- predict(rf_model, newdata = test_data)

conf_matrix_rf <- confusionMatrix(rf_predictions, 
                                  test_data$Exited, 
                                  positive = "1")

conf_matrix_rf
```

###Logistic Regression
Emphasis on interpretation
```{r}
glm_model = glm(Exited ~ ., 
                 data = train_data, 
                 family = binomial(link = "logit"))

summary(glm_model)
```

```{r}
## Evaluation of Logistic Regression Model on Test Set

# 1. Predict probabilities on the test data
# type="response" gives the probability of the positive class (Exited=1)
glm_prob = predict(glm_model, newdata = test_data, type = "response")

# 2. Convert probabilities to class predictions (using default threshold of 0.5)
# This step is standard for generating the confusion matrix.
glm_prediction = factor(ifelse(glm_prob > 0.5, 1, 0), levels = c("0", "1"))

# 3. Create the Confusion Matrix
conf_matrix_glm = confusionMatrix(glm_prediction, 
                                  test_data$Exited, 
                                  positive = "1")

# 4. Print the results
conf_matrix_glm
```


```{r}
## Model Comparison: Reduced Feature Set

# Define the reduced formula based on feature importance analysis:
# We will exclude the dummy variables and keep the original factors/numerics:
reduced_formula = Exited ~ Age + Balance + NumOfProducts + CreditScore + EstimatedSalary + IsActiveMember + Gender

# 1. Train Reduced Random Forest
rf_reduced = randomForest(
  reduced_formula, 
  data = train_data, 
  ntree = 500,
  importance = TRUE
)

# 2. Evaluate Reduced Random Forest
rf_reduced_pred = predict(rf_reduced, newdata = test_data)
conf_matrix_rf_reduced = confusionMatrix(rf_reduced_pred, 
                                        test_data$Exited, 
                                        positive = "1")


# 3. Train Reduced Logistic Regression
glm_reduced = glm(reduced_formula, 
                 data = train_data, 
                 family = binomial(link = "logit"))

# 4. Evaluate Reduced Logistic Regression
glm_reduced_prob = predict(glm_reduced, newdata = test_data, type = "response")
glm_reduced_pred = factor(ifelse(glm_reduced_prob > 0.5, 1, 0), levels = c("0", "1"))
conf_matrix_glm_reduced = confusionMatrix(glm_reduced_pred, 
                                        test_data$Exited, 
                                        positive = "1")
```
```{r}
## Final Performance Comparison

# Function to extract key metrics (You can define this in your Rmd setup)
extract_metrics = function(conf_matrix, model_name) {
  data.frame(
    Model = model_name,
    Accuracy = conf_matrix$overall['Accuracy'],
    Sensitivity = conf_matrix$byClass['Sensitivity'],
    Specificity = conf_matrix$byClass['Specificity'],
    F1_Score = conf_matrix$byClass['F1']
  )
}

# Combine all results
comparison_table = rbind(
  extract_metrics(conf_matrix_rf, "RF_Full"),
  extract_metrics(conf_matrix_rf_reduced, "RF_Reduced"),
  extract_metrics(conf_matrix_glm, "GLM_Full"),
  extract_metrics(conf_matrix_glm_reduced, "GLM_Reduced")
)

# Print the final comparison table
print(comparison_table)
```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```
